### 2019-11-15 main_text/main_text.tex
RP edits plus few minor edits by KL

	% Title must be 250 characters or less.
	\begin{flushleft}
		{\Large
			\textbf\newline{Theoretical investigation of a genetic switch for metabolic adaptation} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
		}
		\newline
		% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
		\\
		Kathrin S. Laxhuber\textsuperscript{1},
		Muir J. Morrison\textsuperscript{2},
		Rob Phillips\textsuperscript{2,3$\ast$}
		\\
		\bigskip
		\textbf{1} Department of Chemistry and Applied Biosciences, ETH Zurich, 8093 Zurich, Switzerland
		\\
		\textbf{2} Department of Physics, California Institute of Technology, Pasadena, CA, USA
		\\
		\textbf{3} Department of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
		\\
		\bigskip

@@

	\end{flushleft}
	% Please keep the abstract below 300 words
	\section*{Abstract}
	Membrane transporters carry important molecules through the cell membrane
	and, from a resource standpoint, should only be produced when necessary. The
	expression of membrane transporters in metabolic pathways is often
	upregulated by the transporter substrate. In \emph{E. coli}, such systems
	include for example the \emph{lacY}, \emph{araFGH}, and \emph{xylFGH} genes,
	which encode for lactose, arabinose, and xylose transporters, respectively.
	As a case study of a minimal system, we build a generizable physical model of the \emph{xapABR} genetic
	circuit, which features a regulatory feedback loop through membrane
	transport (positive feedback) and enzymatic degradation (negative feedback)
	of an inducer. Dynamical systems analysis and stochastic simulations show
	that the membrane transport makes the model system bistable in certain
	parameter regimes. Thus, it serves as a genetic “on-off” switch, enabling
	the system to only produce a set of metabolic enzymes when the corresponding
	metabolite is present in large amounts. We find that the negative feedback
	from the degradation enzyme does not significantly disturb the positive
	feedback from the membrane transporter. We investigate hysteresis in the
	switching and discuss the role of cooperativity and multiple binding sites
	in the model circuit. Fundamentally, this work explores how a stable genetic
	switch for a set of enzymes is obtained from transcriptional auto-activation
	of a membrane transporter through its substrate.

@@

	\section*{Introduction} \todo{remove images \& todonotes everywhere for submission}
	Genetic regulatory circuits are fundamental building blocks of functioning
	cells and organisms. One abundant class of these circuits are genetic
	switches. Although their construction and function may differ,
	their common feature is bistability: their output gene expression
	will flow to and remain at one of two steady-state levels.
	The distribution of
	gene expression in a cell culture can then be bimodal. This is not to be
	confused with mere stochastic bimodality, where the system is not stable,
	and the gene expression in each cell can fluctuate between the two levels.

	One classic example of a genetic switch is a system where two repressor
	proteins each regulate the transcription of the
	other~\cite{Jacob1961,Gardner2000} (illustrated schematically in
	Fig~\ref{fig1:switches}). Here, one stable state is high expression of
	the first protein and low expression of the second,
	and the second stable state is the opposite. This
	switch enables the system to have a memory: if something induces expression
	of either one of the proteins, the system will remain in this state until a
	significant perturbation occurs. Another well known and even simpler example
	is an auto-activating circuit in which a protein activates its own
	transcription~\cite{Wolf1998}. This gives the system an ``on-off'' switch.

@@

	Through physical and mathematical modeling, we investigate a more complex
	switch system where the bistability is due, as we will show, to a
	membrane transport protein. Such a switch is common for
	metabolic processes in biology, for reasons discussed below.
	Existing models in the literature tend towards one of two extremes:
	either highly detailed descriptions of specific, complicated systems (e.g.,~\cite{Wong1997}),
	or Hill function descriptions that coarse-grain all complexity
	into a few parameters with inscrutable microscopic physical meaning.
	We aim for a middle ground in this work.
	We seek an intuitive understanding through a simple model of a minimal system,
	with only the essential components and interactions for the questions we pose.
	Yet we still model these components explicitly and
	discuss the necessary model complexity for a physically correct model.

	The key feature of the type of system we investigate is the indirect activation of
	the transporter gene by the transporter substrate, leading to positive
	feedback similar to the aforementioned ``on-off'' switch. An example for
	such an architecture is the \emph{lac} operon, where lactose indirectly
	activates the expression of lactose permease. Other examples in \emph{E.
		coli} include the \emph{araFGH} and \emph{xylFGH} operons, which contain genes for
	arabinose and xylose transporters, respectively. For \emph{lac} and
	\emph{araFGH}, bistability has indeed been observed and attributed to such a
	positive feedback
	loop~\cite{Novick1957,Santillan2008,Ozbudak2004,Narang2008,Choi2008,Fritz2014,Jenkins2017,Siegele1997}.
	A eukaryotic example is the glucose transporter GLUT-2 in liver and
	\textbeta-cells~\cite{Bae2010,Tiedge1991}, though this system is much more
	complex than the following analysis.

	It is quite conceivable that this auto-activation process is common to many
	substances that a cell would want to accumulate. Such a switch enables the
	cell to sense and respond to its environment: if the substrate enters the
	cell, it activates the production of membrane transporters. The cell then
	starts accumulating the substrate, thereby ``testing'' the substrate's
	presence in the extracellular environment. If there is enough, the
	expression stabilizes at an ``on'' point and the cell has, in a short-term
	sense, adapted. When there is not enough substrate, the operon, which
	often encodes for a whole set of enzymes for this one metabolite, switches
	``off'' again. Such a mechanism could be involved in various cases of
	short-term adaptation.

@@

	\section*{Model}
	\subsection*{Step by step modeling of the system}
	In this section, we present our model of the \emph{xapAB} genetic circuit.
	Fig~\ref{fig2:model} shows an overview of this model.
	The qualitative picture of the circuit switching its state is as follows:
	\begin{itemize}
		\item In the initial absense of XapB, small amounts of xanthosine permeate\todo{``permeate'' wouldn't include active transport (NupC \& NupG), would it? MM: I think it's sufficiently vague to capture all this, details are too subtle to worry about here} into and out of the cell (discussed in more detail below).
		\item XapR is induced by xanthosine.
		\item The induced XapR binds to the \emph{xapAB} promoter, leading to transcription.
		\item From this mRNA transcript, translation produces the two proteins XapA and XapB.
		\item XapB actively transports much larger amounts of xanthosine into the cell and XapA degrades xanthosine.
		\item Production of more XapA \& XapB is balanced by decay or dilution through cell divisions.
	\end{itemize}
	Because xanthosine induces the transcription factor XapR, we have positive
	and negative feedback loops due to XapB and XapA, respectively. The remainder
	of this subsection discusses each of the above steps in detail, leading us
	to a set of two coupled ODEs. More in-depth explanations can be found
	in~\nameref{S1_Text}.

@@

	\paragraph*{Induction of XapR.} 
	We treat dimers as the only form of XapR that appears in the cell. Each
	dimer can bind two xanthosine molecules. The Monod-Wyman-Changeux (MWC)
	model is used to describe the fraction of XapR dimers in the active state,
	which has the form
	\begin{eqnarray}
	\label{eq:MWC}
	\n{[XapR]_A} = \n{[XapR]_{tot}} \frac{\left(1 + \frac{\n{[x]}}{K_{\n{xA}}}\right)^2}{\left(1 + \frac{\n{[x]}}{K_{\n{xA}}}\right)^2 + e^{\beta \Delta E_{\n{x}}} \left(1+\frac{\n{[x]}}{K_{\n{xA}}} \frac{K_{\n{xA}}}{K_{\n{xI}}}\right)^2}.
	\end{eqnarray}
	Here, $\n{[x]}$ is the xanthosine concentration, and $\n{[XapR]_A}$ and
	$\n{[XapR]_{tot}}$ denote the concentration of active and total XapR dimers,
	respectively. Furthermore, $K_{\n{xI}}$ and $K_{\n{xA}}$ are the
	dissociation constants of xanthosine to the inactive and the active XapR
	dimer, respectively, and $\Delta E_{\n{x}}$ stands for the energy difference
	between the inactive and the active state. We expect $\Delta E_{\n{x}} > 0$
	and $K_{\n{xA}} < K_{\n{xI}}$ for inducible activation. This corresponds to
	XapR being mainly inactive in the absence of xanthosine and becoming mostly
	active at high concentrations of xanthosine. A detailed discussion of the
	MWC model can be found in~\cite{Marzen2013}.

@@

	We consider only the state where both XapR binding sites are occupied as
	active and all other states as inactive, meaning they have transcription
	rate equal zero. Experiments show that the expression becomes very weak when one of
	the XapR binding sites is removed from the promoter, suggesting that this
	simplification is reasonable~\cite{Chure2019}. Furthermore, we find that in
	the bistable parameter range, considering the single occupancy states as
	active instead has almost no influence on the results (see
	also~\nameref{S1_Text}).

@@

	\paragraph*{Translation.} 
	The next step in our modeling progression is translation. As a
	simplification, we write $\n{[p]}=\n{[XapA]=[XapB]}$ for the general protein
	concentration. This assumes that the rates of transcription, mRNA decay,
	translation, and protein decay are the same for both proteins, which, as
	discussed in \nameref{S1_Text}, does not have a significant influence on the
	results. We write the following rate equation for the protein concentration,
	where $r_{\n{p}}$ denotes the translation rate, $\gamma_{\n{p}}$ the protein
	decay rate, and $\langle \n{[mRNA]} \rangle$ the mean mRNA concentration:
	\begin{eqnarray}
	\label{eq:pbasic}
	\dd{\n{[p]}}{t} = r_{\n{p}} \langle \n{[mRNA]} \rangle - \gamma_{\n{p}} \n{[p]}.
	\end{eqnarray}

@@

	\todo{MM: do these paragraphs on xanthosine dynamics have an appropriate level of detail now, or too much, or not enough?}
	There are two significant mechanisms for transport of xanthosine across
	the cell membrane. In the induced system, the main transporter is XapB,
	whereas in the uninduced system, there is almost no XapB. Instead,
	xanthosine can enter the cell through the two nucleoside transporters
	NupC and NupG, which have a very low affinity for
	xanthosine~\cite{Norholm2001}. All these transporters, XapB, NupC, and
	NupG, are powered by the proton gradient across the
	membrane~\cite{Norholm2001}, which is why we assume their kinetic scheme
	to be similar to that of the \emph{lac} permease (as it is described
	in~\cite{Kaback2015}). There can be import and export of xanthosine, and
	which one dominates depends on the proton and xanthosine concentrations
	on the two sides of the membrane. In both cases, a proton and a
	substrate need to bind to the transporter on one side of the membrane
	and detach from it on the other side before the empty transporter moves
	back to the other side again. We refer the reader to~\nameref{S1_Text}
	for a detailed description of the transport and its modeling, and just
	state the result here. We model influx and efflux separately. For XapB,
	we use Michaelis-Menten kinetics with parameters $k_{\n{b, i}}$,
	$K_{\n{b, i}}$ for influx and $k_{\n{b, e}}, K_{\n{b,e}}$ for efflux.
	For the Nup transporters, we can linearize because of their low affinity
	for xanthosine and are left with the rate parameters $k_{\n{nup}}$ for
	influx and $\xi k_{\n{nup}}$ for efflux.

	After transport into the cell, XapA degrades xanthosine. We model this using
	standard Michaelis-Menten kinetics, with parameters $k_{\n{a}}, K_{\n{a}}$
	(corresponding to turnover rate and Michaelis constant, respectively). Transport and degradation then leads to the xanthosine rate equation
	\begin{eqnarray}
	\dd{\n{[x]}}{t} = \biggl(\underbrace{k_{\n{b,i}} \frac{c}{K_{\n{b,i}} + c} - k_{\n{b,e}} \frac{\n{[x]}}{K_{\n{b,e}} + \n{[x]}}}_{\n{XapB}} - \underbrace{k_{\n{a}} \frac{\n{[x]}}{K_{\n{a}} + \n{[x]}}}_{\n{XapA}}\biggr) \n{[p]} + \underbrace{k_{\n{nup}} \left(c- \xi \n{[x]}\right)}_{\n{NupC\ \& \ NupG}}.
	\end{eqnarray}
	Recall that $\n{[x]}$ is the intracellular xanthosine concentration, while
	$c$ denotes the extracellular. Because $k_{\n{b,i}} > k_{\n{b,e}}$ and
	$K_{\n{b,i}} < K_{\n{b,e}}$, influx dominates at low intracellular
	xanthosine concentrations. At much higher intra- than extracellular
	xanthosine concentrations, the efflux term takes over. More details on the
	aforementioned steps and a discussion of passive diffusion can be found
	in~\nameref{S1_Text}.

@@

	Very little is known about the \emph{xap} system, and thus, there are almost no
	measured values for the free parameters. Nevertheless, we were able to
	estimate a reasonable range by using values from similar, well studied
	systems and by finding physical constraints or relations between parameters.
	The results of these estimates are shown in Table~\ref{table1:nondim}. They
	are based on a choice of $\gamma_{\n{p}} = 5 \cdot 10^{-4} \unit{s^{-1}}$
	and $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$. A detailed derivation can be
	found in~\nameref{S1_Text}.

@@

	\paragraph*{From a 3D to a 2D system.} 
	Fig~\subref{fig4:bistability}{A} shows the 3D phase portrait for a
	representative set of parameters (shown in Table~\ref{table1:nondim}),
	whose choice is explained below. The plot looks rather complicated at first
	but can be understood intuitively. The three surfaces are the nullcline surfaces and
	the gray lines point in the direction in which the dynamical system moves at
	each point. The surfaces intersect in three points, which are the steady-state
	solutions of the dynamical system. For this choice of parameters, the system
	first flows towards the mRNA nullcline, then it moves along that surface to
	the intersection with the protein nullcline, and lastly, it moves along that
	intersection line to one of the three intersection points of all three
	surfaces.

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Bistability.eps}
		\caption{{\bf Phase portraits showing bistability.}
			3D and 2D phase portraits for one set of parameters that leads to
			bistability. The parameter values are listed in
			Table~\ref{table1:nondim}. Note that all the concentrations
			($\n{[m]_a}, \n{[p]_a}, \n{[x]_a}$) are measured in units of
			$K_{\n{a}} = 5 \cdot 10^{4} \unit{nM}$. The surfaces in (A) and the
			lines in (B) are the nullclines of the state variables, and their
			intersection points, marked in red in (B), are the steady-state
			solutions of the system. The region shaded in gray in (B) leads to
			negative concentrations and is unphysical. A vector plot of (B) that
			also shows the magnitude of flow at each point can be found
			in~\nameref{S1_Text}.}
		\label{fig4:bistability}
	\end{figure}

	It is important to point out that, for a different set of parameters, the
	dynamics can be quite different. There are, for example, scenarios where the
	xanthosine kinetics are roughly as fast as the mRNA kinetics and the
	dynamics unfolds in two steps: first to the intersection of the mRNA and the
	xanthosine nullcline, then along that line to the protein nullcline and
	thereby to a fixed point.

@@

	As already mentioned, we are working with one specific set of parameters
	here and we will now explain this choice of values. Firstly, they were
	picked such that they are well within the parameter range that was estimated beforehand (see Table~\ref{table1:nondim}). Secondly, we
	chose parameters that allow clear bistability in the phase portraits as well as in the
	stochastic simulations (see later), which, of course, is not the case for any possible choice of parameters. Thirdly, by the corresponding choice of
	parameters it was ensured that the mRNA number per cell at the
	``switch-point'' is around 1: this is large enough to enable the system to
	clearly resolve the two stable fixed points (as we will see from the
	stochastic simulations later on), but is low enough to lead to mean mRNA
	numbers that are very reasonable (see~\cite{Milo2016} for the average mRNA
	numbers in cells). The protein and xanthosine
	concentrations followed from this, but with some variation in the parameters
	they could still be tuned to a certain extent.

@@

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/VaryingC.eps}
		\caption{{\bf Phase portraits for different extracellular xanthosine concentrations.}
			All parameters but $\n{[c]_a}$ are as presented in
			Table~\ref{table1:nondim}. The extracellular xanthosine
			concentration in these plots is $\n{[c]_a}=10^{1}$ in (A) and
			$\n{[c]_a}= 5 \cdot 10^{2}$ in (B) (recall that $\n{[c]_{a}} \defeq
			\frac{c}{K_{\n{a}}}$ with $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$,
			so $\n{[c]_a}$ is dimensionless).
			Tuning $\n{[c]_a}$ moves the orange line (xanthosine nullcline) up
			or down, but the blue curve (mRNA nullcline) is unchanged (see
			also~\nameref{S1_Text}). It can clearly be seen that in (A) there is
			only the lower fixed point (fixed point number 1), whereas in (B)
			there is only the upper one (fixed point number 3). In between lies
			the bistable case that was shown in Fig~\ref{fig4:bistability}.}
		\label{fig5:extraxanth}
	\end{figure}

@@

	\paragraph*{The role of cooperativity.} 
	The model has two cooperatively interacting binding sites for XapR on the
	\emph{xapAB} promoter and two cooperative binding sites on XapR for
	xanthosine. An interesting question to pose is whether the cooperativity is
	a necessary feature for bistability. This question is motivated by the importance of
	cooperativity in ``typical'' genetic switches\cite{Gardner2000,Cherry2000}.

>>>>

	% Title must be 250 characters or less.
	\begin{flushleft}
		{\Large
			\textbf\newline{Theoretical investigation of a genetic switch for metabolic adaptation} % Please use "sentence case" for title and headings (capitalize only the first word in a title (or heading), the first word in a subtitle (or subheading), and any proper nouns).
		}
		\newline
		% Insert author names, affiliations and corresponding author email (do not include titles, positions, or degrees).
		\\
		Kathrin S. Laxhuber\textsuperscript{1},
		Muir J. Morrison\textsuperscript{2},
		Rob Phillips\textsuperscript{2,3$\ast$}
		\\
		\bigskip
		\textbf{1} Department of Chemistry and Applied Biosciences, ETH Zurich, 8093 Zurich, Switzerland
		\\
		\textbf{2} Department of Physics, California Institute of Technology, Pasadena, CA, USA
		\\
		\textbf{3} Division of Biology and Biological Engineering, California Institute of Technology, Pasadena, CA, USA
		\\
		\bigskip

@@

	\end{flushleft}
	% Please keep the abstract below 300 words
	\section*{Abstract}
	Membrane transporters carry key metabolites through the cell membrane
	and, from a resource standpoint, are hypothesized to be produced when necessary. The
	expression of membrane transporters in metabolic pathways is often
	upregulated by the transporter substrate. In \emph{E. coli}, such systems
	include for example the \emph{lacY}, \emph{araFGH}, and \emph{xylFGH} genes,
	which encode for lactose, arabinose, and xylose transporters, respectively.
	As a case study of a minimal system, we build a
	generalizable physical model of the \emph{xapABR} genetic
	circuit, which features a regulatory feedback loop through membrane
	transport (positive feedback) and enzymatic degradation (negative feedback)
	of an inducer. Dynamical systems analysis and stochastic simulations show
	that the membrane transport makes the model system bistable in certain
	parameter regimes. Thus, it serves as a genetic ``on-off'' switch, enabling
	the system to only produce a set of metabolic enzymes when the corresponding
	metabolite is present in large amounts. We find that the negative feedback
	from the degradation enzyme does not significantly disturb the positive
	feedback from the membrane transporter. We investigate hysteresis in the
	switching and discuss the role of cooperativity and multiple binding sites
	in the model circuit. Fundamentally, this work explores how a stable genetic
	switch for a set of enzymes is obtained from transcriptional auto-activation
	of a membrane transporter through its substrate.

@@

	\section*{Introduction} \todo{remove images, todonotes, and funding for submission}
	Genetic regulatory circuits are fundamental building blocks of functioning
	cells and organisms. One abundant class of these circuits are genetic
	switches. Although their construction and function may differ,
	their common feature is bistability: their output gene expression
	will flow to and remain at one of two steady-state levels.
	The distribution of
	gene expression in a cell culture can then be bimodal. This is not to be
	confused with mere stochastic bimodality, where the system is not stable,
	and the gene expression in each cell can fluctuate between the two levels.

	One classic example of a genetic switch is a system where two repressor
	proteins each regulate the transcription of the
	other~\cite{Jacob1961,Gardner2000} (illustrated schematically in
	Fig~\ref{fig1:switches}). Here, one stable state is high expression of
	the first protein and low expression of the second,
	and the second stable state is the opposite. This
	switch enables the system to have a memory: if something induces expression
	of either one of the proteins, the system will remain in this state until a
	significant perturbation occurs. Another well-known and even simpler example
	is an auto-activating circuit in which a protein activates its own
	transcription~\cite{Wolf1998}. This gives the system an ``on-off'' switch.

@@

	Through physical and mathematical modeling, we investigate a more complex
	switch system where the bistability is due, as we will show, to a
	membrane transport protein. Such a switch is common for
	metabolic processes in biology, for reasons discussed below.
	Existing models in the literature tend towards one of two extremes:
	either highly detailed descriptions of specific, complicated networks (e.g.,~\cite{Wong1997}),
	or Hill function descriptions that coarse-grain all complexity
	into a few parameters with inscrutable microscopic physical meaning.
	We aim for a middle ground in this work.
	We seek an intuitive understanding through a simple model of a minimal system,
	with only the essential components and interactions for the questions we pose.
	Yet we still model these components explicitly and
	discuss the necessary model complexity for a physically correct model.

	The key feature of the type of system we investigate is the indirect activation of
	the transporter gene by the transporter substrate, leading to positive
	feedback similar to the aforementioned ``on-off'' switch. An example for
	such an architecture is the \emph{lac} operon, where lactose indirectly
	activates the expression of lactose permease. Other examples in \emph{E.
		coli} include the \emph{araFGH} and \emph{xylFGH} operons, which contain genes for
	arabinose and xylose transporters, respectively. For \emph{lac} and
	\emph{araFGH}, bistability has indeed been observed and attributed to such a
	positive feedback
	loop, for example in the well-known study of Novick and Weiner,
	among other works~\cite{Novick1957,Santillan2008,Ozbudak2004,Narang2008,Choi2008,Fritz2014,Jenkins2017,Siegele1997}.
	A eukaryotic example is the glucose transporter GLUT-2 in liver and
	\textbeta-cells~\cite{Bae2010,Tiedge1991}, though this system is much more
	complex than the following analysis.

	It is quite conceivable that this auto-activation process is common to many
	substances that a cell would want to accumulate. Such a switch enables the
	cell to sense and respond to its environment: if the substrate enters the
	cell, it activates the production of membrane transporters. The cell then
	starts accumulating the substrate, thereby ``testing'' the substrate's
	presence in the extracellular environment. If there is enough, the
	expression stabilizes at an ``on'' state and the cell has, in a short-term
	sense, adapted. When there is not enough substrate, the operon, which
	often encodes for a whole set of enzymes for this one metabolite, switches
	``off'' again. Such a mechanism could be involved in various cases of
	short-term adaptation.

@@

	\section*{Model}
	\subsection*{Step by step modeling of the system}
	In this section, we present our model of the \emph{xapAB} genetic circuit.
	Fig~\ref{fig2:model} shows an overview of this model.
	The qualitative picture of the circuit switching its state is as follows:
	\begin{itemize}
		\item In the initial absense of XapB, small amounts of xanthosine permeate
				into and out of the cell (discussed in more detail below).
		\item XapR is induced by xanthosine.
		\item The induced XapR binds to the \emph{xapAB} promoter, leading to transcription.
		\item From this mRNA transcript, translation produces the two proteins XapA and XapB.
		\item XapB actively transports much larger amounts of xanthosine into the cell and XapA degrades xanthosine.
		\item Production of more XapA and XapB is balanced by decay or dilution through cell divisions.
	\end{itemize}
	Because xanthosine induces the transcription factor XapR, we have positive
	and negative feedback loops due to XapB and XapA, respectively. The remainder
	of this subsection discusses each of the above steps in detail, leading us
	to a set of two coupled ODEs. More in-depth explanations can be found
	in~\nameref{S1_Text}.

@@

	\paragraph*{Induction of XapR.} 
	We treat dimers as the only form of XapR that appears in the cell. Each
	dimer can bind two xanthosine molecules~\cite{Joergensen1999}. The Monod-Wyman-Changeux (MWC)
	model is used to describe the fraction of XapR dimers in the active state,
	which has the form
	\begin{eqnarray}
	\label{eq:MWC}
	\n{[XapR]_A} = \n{[XapR]_{tot}} \frac{\left(1 + \frac{\n{[x]}}{K_{\n{xA}}}\right)^2}{\left(1 + \frac{\n{[x]}}{K_{\n{xA}}}\right)^2 + e^{\beta \Delta E_{\n{x}}} \left(1+\frac{\n{[x]}}{K_{\n{xA}}} \frac{K_{\n{xA}}}{K_{\n{xI}}}\right)^2}.
	\end{eqnarray}
	A detailed discussion of the
	MWC model can be found in~\cite{Marzen2013}. In Eq~\ref{eq:MWC},
	$\n{[x]}$ is the xanthosine concentration, and $\n{[XapR]_A}$ and
	$\n{[XapR]_{tot}}$ denote the concentration of active and total XapR dimers,
	respectively. Furthermore, $K_{\n{xI}}$ and $K_{\n{xA}}$ are the
	dissociation constants of xanthosine to the inactive and the active XapR
	dimer, respectively, and $\Delta E_{\n{x}}$ stands for the energy difference
	between the inactive and the active states of the protein. We expect $\Delta E_{\n{x}} > 0$
	and $K_{\n{xA}} < K_{\n{xI}}$ for inducible activation. This corresponds to
	XapR being mainly inactive in the absence of xanthosine and becoming mostly
	active at high concentrations of xanthosine.

@@

	We consider only the state where both XapR binding sites are occupied as
	active and all other states as inactive, meaning they have transcription
	rate equal to zero. Experiments show that the expression becomes very weak when one of
	the XapR binding sites is removed from the promoter, suggesting that this
	simplification is reasonable~\cite{Chure2019}. Furthermore, we find that in
	the bistable parameter range, considering the single occupancy states as
	active instead has almost no influence on the results (see
	also~\nameref{S1_Text}).

@@

	\paragraph*{Translation.} 
	The next step in our modeling progression is translation. As a
	simplification, we write $\n{[p]}=\n{[XapA]=[XapB]}$ for the general protein
	concentration. This assumes that the rates of transcription, mRNA decay,
	translation, and protein decay are the same for both proteins, which, as
	discussed in \nameref{S1_Text}, does not have a significant influence on the
	results. We write the following rate equation for the protein concentration,
	where $r_{\n{p}}$ denotes the translation rate, $\gamma_{\n{p}}$ the protein
	decay rate, and $\langle \n{[m]} \rangle$ the mean mRNA concentration:
	\begin{eqnarray}
	\label{eq:pbasic}
	\dd{\n{[p]}}{t} = r_{\n{p}} \langle \n{[m]} \rangle - \gamma_{\n{p}} \n{[p]}.
	\end{eqnarray}

@@

	There are two significant mechanisms for transport of xanthosine across
	the cell membrane. In the induced system, the main transporter is XapB,
	whereas in the uninduced system, there is almost no XapB. Instead,
	xanthosine can enter the cell through the two nucleoside transporters
	NupC and NupG, which have a very low affinity for
	xanthosine~\cite{Norholm2001}. All these transporters, XapB, NupC, and
	NupG, are powered by the proton gradient across the
	membrane~\cite{Norholm2001}, which is why we assume their kinetic scheme
	to be similar to that of the \emph{lac} permease (as it is described
	in~\cite{Kaback2015}). There can be import and export of xanthosine, and
	which one dominates depends on the proton and xanthosine concentrations
	on the two sides of the membrane. In both cases, a proton and a
	substrate need to bind to the transporter on one side of the membrane
	and detach from it on the other side before the empty transporter moves
	back to the other side again. We refer the reader to~\nameref{S1_Text}
	for a detailed description of the transport and its modeling, and just
	state the result here. We model influx and efflux separately. For XapB,
	we use Michaelis-Menten kinetics with parameters $k_{\n{b, i}}$,
	$K_{\n{b, i}}$ for influx and $k_{\n{b, e}}, K_{\n{b,e}}$ for efflux.
	For the Nup transporters, we also use Michaelis-Menten kinetics but can linearize the Michaelis-Menten term $k_{\n{cat}} \frac{\n{[x]}}{K_{\n{M}} + \n{[x]}} \approx \tilde{k} \n{[x]}$ because of the transporter's low affinity for xanthosine (i.e. $K_{\n{M}} \gg \n{[x]}$). For the rate parameters $\tilde{k}$, we write $k_{\n{nup}}$ for
	influx and $\xi k_{\n{nup}}$ for efflux.

	After transport into the cell, XapA degrades xanthosine. We model this using
	standard Michaelis-Menten kinetics, with parameters $k_{\n{a}}, K_{\n{a}}$
	(corresponding to turnover rate and Michaelis constant, respectively). Transport and degradation then leads to the xanthosine rate equation
	\begin{eqnarray}
	\dd{\n{[x]}}{t} = \biggl(\underbrace{k_{\n{b,i}} \frac{c}{K_{\n{b,i}} + c} - k_{\n{b,e}} \frac{\n{[x]}}{K_{\n{b,e}} + \n{[x]}}}_{\n{XapB}} - \underbrace{k_{\n{a}} \frac{\n{[x]}}{K_{\n{a}} + \n{[x]}}}_{\n{XapA}}\biggr) \n{[p]} + \underbrace{k_{\n{nup}} \left(c- \xi \n{[x]}\right)}_{\n{NupC\ \& \ NupG}}.
	\end{eqnarray}
	Recall that $\n{[x]}$ is the intracellular xanthosine concentration, while
	$c$ denotes the extracellular concentration. Because $k_{\n{b,i}} > k_{\n{b,e}}$ and
	$K_{\n{b,i}} < K_{\n{b,e}}$, influx dominates at low intracellular
	xanthosine concentrations. At much higher intra- than extracellular
	xanthosine concentrations, the efflux term takes over. More details on the
	aforementioned steps and a discussion of passive diffusion can be found
	in~\nameref{S1_Text}.

@@

	Very little is known about the \emph{xap} system, and thus, there are almost no
	measured values for the free parameters. Nevertheless, we were able to
	estimate a reasonable range by using values from similar, well studied
	systems and by exploiting physical constraints or relations between parameters.
	The results of these estimates are shown in Table~\ref{table1:nondim}. They
	are based on a choice of $\gamma_{\n{p}} = 5 \cdot 10^{-4} \unit{s^{-1}}$
	and $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$. A detailed derivation can be
	found in~\nameref{S1_Text}.

@@

	\paragraph*{From a 3D to a 2D system.} 
	Fig~\subref{fig4:bistability}{A} shows the 3D phase portrait for a
	representative set of parameters (shown in Table~\ref{table1:nondim}),
	whose choice is explained below. The plot looks rather complicated at first
	but can be understood intuitively. The three surfaces are the nullcline surfaces and
	the gray lines point in the direction in which the dynamical system moves at
	each point. The surfaces intersect in three points, which are the steady-state
	solutions of the dynamical system. For this choice of parameters, the system
	first flows towards the mRNA nullcline (independent of the initial condition),
	then it moves along that surface to
	the intersection with the protein nullcline, and lastly, it moves along that
	intersection line to one of the three intersection points of all three
	surfaces.

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/Bistability.eps}
		\caption{{\bf Phase portraits showing bistability.}
			3D and 2D phase portraits for one set of parameters that leads to
			bistability. The parameter values are listed in
			Table~\ref{table1:nondim}. Note that all the concentrations
			($\n{[m]_a}, \n{[p]_a}, \n{[x]_a}$) are measured in units of
			$K_{\n{a}} = 5 \cdot 10^{4} \unit{nM}$. The surfaces in (A) and the
			curves in (B) are the nullclines of the state variables, and their
			intersection points, marked in red in (B), are the steady-state
			solutions of the system. The region shaded in gray in (B) leads to
			negative concentrations and is unphysical. A vector plot of (B) that
			also shows the magnitude of flow at each point can be found
			in~\nameref{S1_Text}.}
		\label{fig4:bistability}
	\end{figure}

	It is important to point out that, for a different set of parameters, the
	dynamics can be quite different. There are, for example, scenarios where the
	xanthosine kinetics are roughly as fast as the mRNA kinetics and the
	dynamics unfolds in two steps: first to the intersection of the mRNA and the
	xanthosine nullcline, then along that curve to the protein nullcline and
	thereby to a fixed point.

@@

	As already mentioned, we are working with one specific set of parameters
	here and we will now explain this choice of values. Firstly, they were
	picked roughly in the middle of the range that was estimated beforehand for this parameter (see Table~\ref{table1:nondim} and~\nameref{S1_Text}). Secondly, we
	chose parameters that allow clear bistability in the phase portraits as well as in the
	stochastic simulations (see later), which, of course, is not the case for any possible choice of parameters. Thirdly, by the corresponding choice of
	parameters it was ensured that the mRNA number per cell at the
	``switch-point'' is around 1: this is large enough to enable the system to
	clearly resolve the two stable fixed points (as we will see from the
	stochastic simulations later on), but is low enough to lead to mean mRNA
	numbers that are very reasonable (see~\cite{Milo2016} for the average mRNA
	numbers in bacterial cells). The protein and xanthosine
	concentrations followed from this, but with some variation in the parameters
	they could still be tuned to a certain extent.

@@

	\begin{figure}%[h!]
		\centering
		\includegraphics{media/VaryingC.eps}
		\caption{{\bf Phase portraits for different extracellular xanthosine concentrations.}
			All parameters but $\n{[c]_a}$ are as presented in
			Table~\ref{table1:nondim}. The extracellular xanthosine
			concentration in these plots is $\n{[c]_a}=10^{1}$ in (A) and
			$\n{[c]_a}= 5 \cdot 10^{2}$ in (B) (recall that $\n{[c]_{a}} \defeq
			\frac{c}{K_{\n{a}}}$ with $K_{\n{a}} = 5 \cdot 10^{-5} \unit{M}$,
			so $\n{[c]_a}$ is dimensionless).
			Tuning $\n{[c]_a}$ moves the orange line (xanthosine nullcline), but the blue curve (mRNA nullcline) is unchanged (see
			also~\nameref{S1_Text}). It can clearly be seen that in (A) there is
			only the lower fixed point (fixed point number 1), whereas in (B)
			there is only the upper one (fixed point number 3). In between lies
			the bistable case that was shown in Fig~\ref{fig4:bistability}.}
		\label{fig5:extraxanth}
	\end{figure}

@@

	\paragraph*{The role of cooperativity.} 
	The model has two cooperatively interacting binding sites for XapR on the
	\emph{xapAB} promoter and two cooperative binding sites on XapR for
	xanthosine. It is interesting to consider whether the cooperativity is
	a necessary feature for bistability. This question is motivated by the importance of
	cooperativity in ``typical'' genetic switches\cite{Gardner2000,Cherry2000}.

	If, as a purely theoretical consideration, we remove either the second
	xanthosine binding site on XapR or the second XapR binding site on the
	promoter, leaving cooperativity in only one component of the system,
	we find that the system still has a bistable parameter regime.
	However, this bistable parameter range is smaller than in the original
	model, which makes the system less stable: small stochastic fluctuations
	in the parameter values can collapse the system to monostability,
	possibly leaving it in the wrong state and without its ability to adapt.
	But only when the second binding site is removed in both places, leaving
	no cooperativity in the system, do we find that it is insufficiently non-linear to produce bistability. An example
	of the three scenarios (only cooperative XapR, only cooperative promoter, no cooperativity) can be seen in Fig~\ref{fig7:coop}. It follows that
	there need to be either two xanthosine binding sites on XapR or two XapR
	binding sites on the promoter (or both) in order to obtain a switch-like
	behavior.
